Universal algebra and the \texorpdfstring{$ \lambda $}{lambda}-calculus

### 2024-05-29 report/src/category-theory.tex
Define the theory of extensions

>>>>

\section{Coends}\label{sec:coends}
This section is based on Section 1.2 of \autocite{riehl}.

In this thesis, we will encounter co-ends a couple of times, so we will introduce them here.
\begin{definition}
  Let $ C, D $ be categories and $ F : \op C \times C \to D $ a functor. We define the \iindex{coend} $ \int^{c : C} F(c, c) $ to be the colimit
  \begin{center}
    \begin{tikzcd}
      \displaystyle\coprod_{\substack{f : C(a, b)}} F(b, a)
        \arrow[r, shift left, "{F(f, \id b)}"]
        \arrow[r, shift right, "{F(\id a, f)}"'] &
      \displaystyle\coprod_{c : C} F(c, c)
        \arrow[r, dashed] &
      \int^C F
    \end{tikzcd}
  \end{center}
\end{definition}

\begin{remark}
  An alternative way to phrase this, is that $ \int^C F : D $ is an object, equipped with arrows $ F(c, c) \to \int^C F $ such that for all $ f: C(a, b) $, the following diagram must commute
  \begin{center}
    \begin{tikzcd}
      F(b, a) \arrow[r, "{F(f, \id a)}"] \arrow[d, "{F(\id b, f)}"] & F(a, a) \arrow[d]\\
      F(b, b) \arrow[r] & \int^C F
    \end{tikzcd}
  \end{center}
  and such that for any other $ G : D $ with the same properties, we have a unique morphism $ \int_C F \to G $, making the triangles commute
  \begin{center}
    \begin{tikzcd}
      F(c, c) \arrow[r] \arrow[rd] & \int_C F \arrow[d]\\
      & G
    \end{tikzcd}
  \end{center}
\end{remark}

\begin{remark}
  Of course, a co-end is actually the dual notion of an end, which can be defined as the equalizer of the diagram above, but with the arrows reversed.
\end{remark}

\begin{remark}
  Left Kan extension can be expressed as a coend:
  \[ \Lan F G(b) = \int^{a : A} D(F a, b) \cdot G a \]
  where $ S \cdot X $ for $ S $ a set and $ X : C $ denotes the `copower'. In most cases, we will have
  \[ S \cdot X = \coprod_{s : S} X. \]
\end{remark}

### 2024-05-29 report/src/paper.tex
Define the theory of extensions

\begin{lemma}
  The category of $ T $-algebras has coproducts.
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

\begin{definition}[Theory of extensions]
  Let $ T $ be an algebraic theory and $ A $ a $ T $-algebra. We can define an algebraic theory $ T_A $ called `the theory of extensions of $ A $' with $ (T_A)_n = T_n + A $. The left injection of the variables $ x_i : T_n $ gives the variables.
  Now, take $ h: (T_n + A)^m $. Sending $ g: T_m $ to $ \varphi(g) := g \bullet h $ gives a $ T $-algebra morphism $ T_m \to T_n + A $ since
  \[ \varphi(f \bullet g) = f \bullet g \bullet h = f \bullet (g_i \bullet h) = f \bullet (\varphi(g_i))_i. \]
  This, together with the injection morphism of $ A $ into $ T_n + A $, gives us a $ T $-algebra morphism from the coproduct: $ T_m + A \to T_n + A $. We especially have a function on sets $ (T_m + A) \times (T_n + A)^m \to T_n + A $, which we will define our substitution to be.
\end{definition}

\begin{lemma}
  $ T_A $ is indeed an algebraic theory.
\end{lemma}
\begin{proof}
  \TODO
\end{proof}

>>>>

The fundamental theorem of the $ \lambda $-calculus that Hyland shows is actually not of the form shown above. To get there, we first need to show that the category of $ T $-algebras for an algebraic theory $ T $ has coproducts, and define the `theory of extensions'.

Let $ \llbracket n \rrbracket $ denote the finite set $ \{ 1, 2, \dots, n \} $. Let \index{$ \F $}$ \F \subseteq \FinSET $ be the full subcategory with $ \F_0 = \{ \llbracket 0 \rrbracket, \llbracket 1 \rrbracket, \dots \} $.

\begin{lemma}
  Let $ T $ be an algebraic theory. The category of $ T $-algebras has coproducts.
\end{lemma}
\begin{proof}
  This is shown in \autocite{algebaric-theories-2010}, in the lemmas leading up to Theorem 4.5.

  Explicitly, we can express the coproduct of algebras, and especially its set, as the following coend (\autocite{Hyland}, Proposition 2.5) (see also Section \ref{sec:coends} for more on coends)
  \[ A + B = \int^{(m, n) : \F \times \F} A^m \times B^n \times T_{m^\prime + n^\prime}, \]
  considering $ T $ as a functor $ T : \F \to \SET $ as a functor (see Remark \ref{rem:abstract-clone-functor}). Note that we do not need the exact definition of $ A + B $ for the rest of this section. Nonetheless, it is interesting to see how it is defined and why this definition works.

  One can think of $ A + B $ consisting of elements $ t \bullet (a + b) $ for $ t: T_{m + n} $, $ a : A^m $ and $ b: B^n $ (writing $ (a + b) $ for $ (a_1, \dots, a_m, b_1, \dots, b_n) $), `substituting' the $ a_i $ and $ b_j $ for the $ x_i $ and $ x_{m + j} $ in $ t $.

  The relations then give `associativity' of this substitution and also assure that reordering or duplicating $ x_i $ and their corresponding $ a_i $ and $ b_j $ do not yield different elements: associating the images on the left and right of
  \[ A^{m^\prime} \times B^{n^\prime} \times T_{m^\prime + n^\prime} \leftarrow A^{m^\prime} \times B^{n^\prime} \times T_{m + n} \rightarrow A^m \times B^n \times T_{m + n} \]
  gives
  \[ (t \bullet ((x_{f(i)})_i + (x_{m + g(j)})_j)) \bullet a + b = t \bullet ((a_{f(i)})_i + (b_{g(j)})_j) \]
  for $ f: \llbracket m \rrbracket \to \llbracket m^\prime \rrbracket $, $ g: \llbracket n \rrbracket \to \llbracket n^\prime \rrbracket $, $ t : T_{m + n} $, $ a : A^{m^\prime} $ and $ b : B^{n^\prime} $.

  Then it becomes clear what the action of $ T $ will be on $ A + B $, although the precise definition has quite a combinatorial nature. For $ s : T_l $, $ t_i : T_{m_i + n_i} $, $ a_i : A^{m_i} $ and $ b_i : B^{n_i} $, define the disjoint embeddings
  \[ d_i : \left\llbracket m_i + n_i \right\rrbracket \cong \left\llbracket m_i \right\rrbracket \sqcup \left\llbracket n_i \right\rrbracket \hookrightarrow \left\llbracket \sum_j m_j \right\rrbracket \sqcup \left\llbracket \sum_j n_j \right\rrbracket \cong \left\llbracket \sum_j m_j + \sum_j n_j \right\rrbracket, \]
  which we will use to make sure that the $ x_j $ in the different $ t_i $ are mapped to distinct variables. Then we can define
  \[ s \bullet (t_i \bullet a_i + b_i) = (s \bullet (t_i \bullet (x_{d_i(j)})_j)_i) \bullet (a_1 + b_1 + \dots + a_l + b_l). \]
  More formally (using the coend injections $ A^{\sum_k m_k} \times B^{\sum_k n_k} \times T_{\sum_k m_k + \sum_k n_k} \to A + B $), this gives functions
  \[ T_l \to (A^{m_1} \times B^{n_1} \times T_{m_1 + n_1} \to (\dots \to (A^{m_l} \times B^{n_l} \times T_{m_l + n_l} \to A + B) \dots )), \]
  commuting with the relations between the different $ (A^m \times B^n \times T_{m + n}) $, which, by repeatedly using the universal property of the coend, then correspond to functions
  \[ T_l \to (A + B \to (\dots \to (A + B \to A + B) \dots )), \]
  or, equivalently, a function
  \[ T_l \times (A + B)^l \to A + B. \]

  We have left and right injections $ A \to A + B $ and $ B \to A + B $, given respectively by the maps $ A^1 \times B^0 \times T_{1 + 0} \to A + B $ and $ A^0 \times B^1 \times T_{0 + 1} \to A + B $:
  \[ a \mapsto x_1 \bullet a \quad \text{and} \quad b \mapsto x_1 \bullet b \]
  and every element $ A + B $ arises by the action of $ t $ on combinations of these embedded elements:
  \[ t \bullet (a + b) = t \bullet ((x_1 \bullet a_i)_i + (x_1 \bullet b_j)_j), \]
  which ultimately can be used to show that $ A + B $ indeed has the universal property of the coproduct.
\end{proof}

\begin{definition}[Theory of extensions]
  Let $ T $ be an algebraic theory and $ A $ a $ T $-algebra. We can define an algebraic theory $ T_A $ called `the theory of extensions of $ A $' with $ (T_A)_n = T_n + A $. The left injection of the variables $ x_i : T_n $ gives the variables.
  Now, take $ h: (T_n + A)^m $. Sending $ g: T_m $ to $ g \bullet h $ gives a $ T $-algebra morphism $ T_m \to T_n + A $. Together with the right injection morphism of $ A $ into $ T_n + A $, this gives us a $ T $-algebra morphism from the coproduct: $ T_m + A \to T_n + A $. In particular, we have a function on sets $ (T_m + A) \times (T_n + A)^m \to T_n + A $, which we will define our substitution to be.
\end{definition}
